A sample article title

### 2019-05-15 bmc_article.tex
updated manuscript

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. [...]

@@

Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF]. We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations.

@@

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeister_sctransform_2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary XX). In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data. This was not the case in the smartseq datasets, which required the additional regressing out of covariates.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure XX). For many clusters, scaling tended to reduce both the average silhouette width (XXX) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure XXX), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure XXX) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure XXX). Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure XXX A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure XXX B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures XX and XX). We focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected. A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figure XX). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure XX). To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.

These main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found Seurat's unstandardized estimate of variance to provide the best results across all normalization methods, followed by deviance.

We next investigated the impact of the number of features selected.

@@

\subsection*{Sub-heading for section}
Text for this sub-heading \ldots
\subsubsection*{Sub-sub heading for section}
Text for this sub-sub-heading \ldots
\paragraph*{Sub-sub-sub heading for section}
Text for this sub-sub-sub-heading \ldots
In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
addition, we examine a common modeling assumption and note the
importance of considering the tails of the extinction time $T_x$ in
studies of escape dynamics.
We will first consider the expected resistant population at $vT_x$ for
some $v>0$, (and temporarily assume $\alpha=0$)

Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}

>>>>

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure 2), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure 2). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) which are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoids most of the subpopulation bias while excluding strong outliers.

@@

Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF]. We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 11). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an imperfect correlation \citep{hafemeister_sctransform_2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features (Supplementary Figure 12A). In the 10X datasets, most methods successfully decreased the correlation with covariates, including the mere scaling of non-normalized data. This was not the case in the smartseq datasets, which required the additional regressing out of covariates.

@@

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure 3 and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure 3). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure 3), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure 3 and Supplementary Figure 13). Interestingly, clusters with the lowest average silhouette width (and which are therefore more difficult to classify) tended to be less well distinguished with sctransform, and instead showed a good difference in non-normalized and/or non-scaled data, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 14B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 15-16). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure 4A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high _deviance_ explained by subpopulations (Figure 4 and Supplementary Figures 15-16). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 17). To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.

The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and Seurat's unstandardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.

@@
